---
title: "practical_exercise_2, Methods 3, 2021, autumn semester"
author: '[FILL IN YOUR NAME]'
date: "[FILL IN THE DATE]"
output: pdf_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, lme4)
```

# Assignment 1: Using mixed effects modelling to model hierarchical data
In this assignment we will be investigating the _politeness_ dataset of Winter and Grawunder (2012) and apply basic methods of multilevel modelling. 

## Dataset
The dataset has been shared on GitHub, so make sure that the csv-file is on your current path. Otherwise you can supply the full path.

```{r}
politeness <- read.csv('politeness.csv') ## read in data
```

# Exercises and objectives
The objectives of the exercises of this assignment are:  
1) Learning to recognize hierarchical structures within datasets and describing them  
2) Creating simple multilevel models and assessing their fitness  
3) Write up a report about the findings of the study  

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below  
REMEMBER: This assignment will be part of your final portfolio

## Exercise 1 - describing the dataset and making some initial plots

### 1) Describe the dataset, such that someone who happened upon this dataset could understand the variables and what they contain  
    i. Also consider whether any of the variables in _politeness_ should be encoded as factors or have the factor encoding removed. Hint: ```?factor```  

```{r}

politeness <- politeness %>%
    filter(!is.na(f0mn))

politeness
```
The dataset consists of seven variables: 
  subject = contains the subject ID,
  gender = F indicating that the subject is female and M indicating that the subject is male.
  scenario = each scenario being a different sentence that the subject had to say
  attitude = indicating whether the sentence was said in a “polite” or “informal” context.
  total_duration = 
  f0mn = the measured voice pitch, where higher values is higher pitch
  hiss_count = 

### 2) Create a new data frame that just contains the subject _F1_ and run two linear models; one that expresses _f0mn_ as dependent on _scenario_ as an integer; and one that expresses _f0mn_ as dependent on _scenario_ encoded as a factor  
    **i. Include the model matrices, $X$ from the General Linear Model, for these two models in your report and describe the different interpretations of _scenario_ that these entail**

```{r}

f1 <- politeness %>%
  filter(subject == "F1") %>%
  mutate(scenario_factor = scenario)

f1$scenario_factor <- as.factor(f1$scenario_factor)
  

lm_s <- lm(f0mn ~ scenario, f1)

lm_f <- lm(f0mn ~ scenario_factor, f1)


summary(lm_s)
summary(lm_f)


```
As we see, we get two very different outputs, especially when looking at the coefficients. When _scenario_ is modeled as integer, the model treats the variable as a continuum, whereas when coded as a factor, _scenario_ is treated categorically.

  **ii. Which coding of _scenario_, as a factor or not, is more fitting?**
  The factorized version is more fitting. Because the numbers in scenario should not be treated as continuum (no floats), but should instead be treated as independent categories, which becomes the case when _scenario_ is factorized.
    
### 3) Make a plot that includes a subplot for each subject that has _scenario_ on the x-axis and _f0mn_ on the y-axis and where points are colour coded according to _attitude_.
    i. Describe the differences between subjects
    
```{r}

politeness$scenario <- as.factor(politeness$scenario)

ggplot(politeness, aes(x = scenario, y = f0mn, color = attitude)) +
  geom_point() +
  facet_wrap(~subject) +
  labs(title = "Subplots for each subject") +
  theme_bw()


```
First of all, we see a difference between female an male subjects, where females have a higher pitch than males in general. Next, the plots gives an indication of having a higher pitch in inpolite context than polite context.
    
## Exercise 2  - comparison of models

For this part, make sure to have `lme4` installed.  
You can install it using `install.packages("lme4")` and load it using `library(lme4)`  
`lmer` is used for multilevel modelling

```{r, eval=FALSE}
mixed.model <- lmer(formula=..., data=...)
example.formula <- formula(dep.variable ~ first.level.variable + (1 | second.level.variable))
```

### 1) Build four models and do some comparisons
  
  **i. a single level model that models _f0mn_ as dependent on _gender_**

```{r}
model_1 <- lm(f0mn ~ gender, data = politeness)
```

  **ii. a two-level model that adds a second level on top of i. where unique intercepts are modelled for each _scenario_**

```{r}
model_2 <- lmer(f0mn ~ gender + (1|scenario), data = politeness)

```

  **iii. a two-level model that only has _subject_ as an intercept**

```{r}
model_3 <- lmer(f0mn ~ gender + (1|subject), data = politeness)

```

  **iv. a two-level model that models intercepts for both _scenario_ and _subject_**

```{r}

model_4 <- lmer(f0mn ~ gender + (1|subject) + (1|scenario), data = politeness)

```

  **v. which of the models has the lowest residual standard deviation, also compare the Akaike Information Criterion `AIC`?**
```{r}

sum(residuals(model_1)^2)
sum(residuals(model_2)^2)
sum(residuals(model_3)^2)
sum(residuals(model_4)^2)

AIC(model_1, model_2, model_3, model_4)

```

  **vi. which of the second-level effects explains the most variance?**
_subject_ seems to explain the most variance compared to _scenario_. The AIC is also lower for the model with _subject_ as the intercept compared to the model with _scenario_ as intercept.

### 2) Why is our single-level model bad?
  **i. create a new data frame that has three variables, _subject_, _gender_ and _f0mn_, where _f0mn_ is the average of all responses of each subject, i.e. averaging across _attitude_ and_scenario_**

```{r}
politeness_2 <- politeness %>%
  select(subject = "subject", gender = "gender", f0mn = "f0mn") %>%
  filter(!is.na(f0mn)) %>%
  group_by(subject, gender) %>%
  summarise(f0mn_mean = mean(f0mn))

politeness_2

```

    
  ii. build a single-level model that models _f0mn_ as dependent on _gender_ using this new dataset
  
```{r}

model_5 <- lm(f0mn_mean ~ gender, data = politeness_2)

summary(model_5)

```
  
  **iii. make Quantile-Quantile plots, comparing theoretical quantiles to the sample quantiles) using `qqnorm` and `qqline` for the new single-level model and compare it to the old single-level model (from 1).i). Which model's residuals ($\epsilon$) fulfil the assumptions of the General Linear Model better?)**

```{r}

qqnorm(residuals(model_1))
qqline(residuals(model_1))


qqnorm(residuals(model_5))
qqline(residuals(model_5))

```

  iv. Also make a quantile-quantile plot for the residuals of the  multilevel model with two intercepts. Does it look alright?
  
```{r}

qqnorm(residuals(model_4))
qqline(residuals(model_4))

```

  
### 3) Plotting the two-intercepts model
  
  i. Create a plot for each subject, (similar to part 3 in Exercise 1), this time also indicating the fitted value for each of the subjects for each for the scenarios (hint use `fixef` to get the "grand effects" for each gender and `ranef` to get the subject- and scenario-specific effects)
  
```{r}

ggplot(politeness, aes(x = scenario, y = f0mn, group = subject)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  facet_wrap(~ subject) +
  labs(title = "Subplots for each subject") +
  theme_bw()

fixef(model_4)
ranef(model_4)


```

    
## Exercise 3 - now with attitude

### 1) Carry on with the model with the two unique intercepts fitted ( _scenario_ and _subject_ ).

  **i. now build a model that has _attitude_ as a main effect besides _gender_**

```{r}

model_6 <- lmer(f0mn ~ gender + attitude + (1|subject) + (1|scenario), data = politeness)
summary(model_6)

```

  **ii. make a separate model that besides the main effects of _attitude_ and _gender_ also include their interaction**

```{r}
model_7 <- lmer(f0mn ~ gender*attitude + (1|subject) + (1|scenario), data = politeness)
summary(model_7)
```

  **iii. describe what the interaction term in the model says about Korean men's pitch when they are polite relative to Korean women's pitch when they are polite (you don't have to judge whether it is interesting)  **

An interaction indicates whether there is some relationsship between the two independent variables - whether the effect of $x_1$ on $y$ depends on the value of $x_2$ and vice versa.

The model indicates that there are some interaction between the two predictors, however, this interaction is not significant.

### 2) Compare the three models (1. gender as a main effect; 2. gender and attitude as main effects; 3. gender and attitude as main effects and the interaction between them. For all three models model unique intercepts for _subject_ and _scenario_) using residual variance, residual standard deviation and AIC.  

### 3)  Choose the model that you think describe the data the best - and write a short report on the main findings based on this model. At least include the following:
  i. describe what the dataset consists of  
  ii. what can you conclude about the effect of gender and attitude on pitch (if anything)?  
  iii. motivate why you would include separate intercepts for subjects and scenarios (if you think they should be included)  
  iv. describe the variance components of the second level (if any)  
  v. include a Quantile-Quantile plot of your chosen model  